\myTitle

### 2024-05-28 tex/body-text-of-paper.tex
More overleaf changes

 In the 1980s, electrophoretically
detectable genetic variation, in the form of allozymes
\citep{ayala1972allozymes,allendorf1981use}, was used to
establish a program of GSI for Chinook salmon,
{\em Oncorhynchus tshawytscha} \citep{milner1982genetic}.  Extensive sampling
revealed that these allozyme markers
exhibited different allele frequencies among major stocks of Chinook salmon on the West Coast.
These allele frequency differences make GSI possible, and
\citet{milner1985genetic} soon showed that proportions of stocks in
the Washington state coastal troll fishery could be estimated by GSI.
Since that time, with the development of novel molecular markers, and, now
with increasing capacity to sequence genomic material, the scope and scale of GSI
has expanded considerably.

@@

The ascendance over the last decade of parentage based tagging (PBT) as a management
tool for Pacific salmon adds another factor to consider when developing a GSI baseline.
Since the first proposal \citep{anderson2005description} to replace or augment the coded-wire tag
programme \citep{nandor2010overview}
with PBT, it has been noted that one of the major advantages of a genetic program for
PBT is that the genetic markers used for PBT could also be useful for GSI (and
vice versa).

Thus, any panel of markers to be used for GSI (or PBT) should also be evaluated on its
utility for PBT (or GSI).  PBT has been remarkably successful in fisheries
management, having largely replaced coded wire tags in large parts of the Snake River and
Columbia River basin (CITATION), and having been employed to dramatically further our
understanding of the genetic inheritance of key traits in salmonids
\citep{abadia2013large,beulke2023distinct}. However, PBT is just one subset of whole
family of statistical genetic methods employing relationship inference to learn about
populations.  For example inference of the full siblings amongst a sample of fish
provides information about the effective number of adults producing offspring
\citep{waples2011inbreeding,wang2023estimating}, which can
be a valuable source of information when sampling of the adults is not possible.

@@

@@

We sampled fish for the reference baselines from 13 locations within California, and one 
within Oregon.
These locations included sites in the two
main tributaries (the Sacramento River and the San Joaquin River) within
California's Central Valley, and the two main tributaries (Klamath and Trinity Rivers) in the Klamath basin in Northern California, and
3 rivers of the California coast (Russian, Eel, and Smith)
(Fig.~\ref{fig:map}, Table~\ref{tab:samples}).
%%%%%%%%%%%%%
\begin{figure}
\newcommand{\mapcap}{\footnotesize  A map of sampling locations represented in the reference 
baseline.  The colored circles aside
each location name represent the different run-timing groups of Chinook salmon
included in the collections from the location.  Color codes are as follows: \Wball~=~Winter run,
\LFball{}~=~Late-fall run, \Fball~=~Fall run, \Sball~=~Spring run. 
Names of each location are followed by the codes used to refer to fish from each location, with
{\tt XXF/S} indicating {\tt XXF} and {\tt XXS} are used to denote fall-run and spring-run fish
from location {\tt XX}.  Abbreviations used in location names as follows: R. = River; Ck. = Creek; H. = Hatchery; NFH = National Fish Hatchery.}
\begin{center}
\includegraphics[width=\columnwidth]{images/map-crop.pdf}
\end{center}
\caption[\mapcap]{\mapcap}
\label{fig:map}
\end{figure}
%%%%%%%%%%%%%%%

@@

Accurate identification of the distinct ecotypes of Chinook salmon in the California Central Valley is of utmost importance for designing and implementing effective conservation and management actions. This has been challenging, given the recent common ancestors of these ecotypes and ongoing migration between subbasins. We describe the first set of genetic markers that identifies the late-fall-run salmon ecotype, which occurs only in the CCV and shares a genomic background with the more common CCV fall-run salmon ecotype. Moreover, we show how the FRH spring-run lineage is easily identifiable through a combination of traditional GSI and the characterization of functional genetic markers in the RoSA. Finally, although previous work has described GSI capabilities that distinguish the natural-spawning CCV spring-run lineages from each other and their fall-run counterparts with moderately high accuracy, we demonstrate near complete accuracy in distinguishing these 'stocks'. Moreover, the few fish that are apparently misidentified (Fig. 3) are likely to be primarily migrants and not true misidentifications. For example, the three fish that were field-characterized as spring-run from Butte Creek, but are genetically identified as winter-run salmon are clearly carry the winter-run genomic background and could not realistically be misidentified on the basis of the genetic data. This marker set and baseline reference dataset also provides unprecedented power for identifying fish from not only the other ESUs in California, but even the individual populations within those ESUs (Fig. 3).
Additionally, previous genetic marker sets for California Chinook salmon had sufficient power for inferring parent-offspring relationships, but only when both parents were sampled and genotyped. This marker set considerably increases the capacity for relationship inference in California salmon, by providing sufficient power for parent-offspring relationship inference, when only one parent was sampled, as well as identifying pairs of full siblings when no parents are sampled. Groups of full siblings larger than two are identified even more accurately than expressed in Figure 4.

>>>>

 In the 1980s, electrophoretically
detectable genetic variation, in the form of allozymes
\citep{ayala1972allozymes,allendorf1981use}, was used to
establish a program of GSI for Chinook salmon,
{\em Oncorhynchus tshawytscha} \citep{milner1982genetic}.  Extensive sampling
revealed that these allozyme markers
exhibited different allele frequencies among major lineages, or stocks, of Chinook salmon on the West Coast.
These allele frequency differences make GSI possible, and
\citet{milner1985genetic} soon showed that proportions of stocks in
the Washington state coastal troll fishery could be estimated by GSI.
Since that time, with the development of novel molecular markers, and, now
with increasing capacity to sequence genomic material, the scope and scale of GSI
has expanded considerably.

@@

The emergence over the last decade of parentage based tagging (PBT) as a management
tool for Pacific salmon adds another factor to consider when developing a GSI baseline.
Since the first proposal \citep{anderson2005description} to replace or augment the coded-wire tag
programme \citep{nandor2010overview}
with PBT, it has been noted that one of the major advantages of a genetic program for
PBT is that the genetic markers used for PBT could also be useful for GSI (and
vice versa).

Thus, any panel of markers to be used for GSI (or PBT) should also be evaluated on its
utility for PBT (or GSI).  PBT has been remarkably successful in fisheries
management, having been used for over a decade in the management of Chinook salmon
and steelhead trout in major basins of the Columbia River \cite{steele2019parentage,horn2023multigeneration}, and having been employed to dramatically further our
understanding of the genetic inheritance of key traits in salmonids
\citep{abadia2013large,beulke2023distinct}. However, PBT is just one subset of a whole
family of statistical genetic methods employing relationship inference to learn about
populations.  For example inference of the full siblings amongst a sample of fish
provides information about the effective number of adults producing offspring
\citep{waples2011inbreeding,wang2023estimating}, which can
be a valuable source of information when sampling of the adults is not possible.

@@

Chinook salmon are the largest of the Pacific salmonids and have historically been the target of extremely high-value and culturally important fisheries. Moreover, because of their high degree of ecotypic variation, they have provided fishery opportunities during many different seasons and geographic locations. However, recent declines in population sizes, from the Yukon River in Arctic north, to the southern extent of their range in California has led to almost complete fishery closures  throughout most of the northeastern Pacific to protect less productive stocks. The co-occurence of fish from relatively productive and relict populations reaches its paragon in California, with the largest remaining ocean fisheries for Chinook salmon targeting the Central Valley fall-run stock that spawns in different tributaries of the same river basin as the highly endangered and phenotypically distinct winter-run stock and the threatened spring-run stocks. In spite of their distinct phenotypes and life history patterns, the four ecotypes of Chinook salmon in the Central Valley are very closely related \cite{clemento2014evaluation} and share recent common ancestry that is independent of ongoing migration between these lineages. As such, GSI has been particularly challenging in the CCV, with at least one of the distinct ecotypes, late-fall-run, unresolvable with previous GSI marker sets \cite{seeb2007development,clemento2014evaluation} and with unsatisfactory power for discriminating the protected spring-run stocks from the harvested fall-run stock.

@@

We sampled fish for the reference baselines from 13 locations within California, and one 
within Oregon. This set of populations and stocks includes all of the previously described lineages of Chinook salmon in the southern portion of the range and most of the populations that are known to have significant genetic differentiation in this region. 
These locations included sites in the two
main tributaries (the Sacramento River and the San Joaquin River) within
California's Central Valley, and the two main tributaries (Klamath and Trinity Rivers) in the Klamath basin in Northern California, and
3 rivers of the California coast (Russian, Eel, and Smith)
(Fig.~\ref{fig:map}, Table~\ref{tab:samples}).
%%%%%%%%%%%%%
\begin{figure}
\newcommand{\mapcap}{\footnotesize  A map of sampling locations represented in the reference 
baseline.  The colored circles aside
each location name represent the different run-timing groups of Chinook salmon
included in the collections from the location.  Color codes are as follows: \Wball~=~Winter run,
\LFball{}~=~Late-fall run, \Fball~=~Fall run, \Sball~=~Spring run. 
Names of each location are followed by the codes used to refer to fish from each location, with
{\tt XXF/S} indicating {\tt XXF} and {\tt XXS} are used to denote fall-run and spring-run fish
from location {\tt XX}.  Abbreviations used in location names as follows: R. = River; Ck. = Creek; H. = Hatchery; NFH = National Fish Hatchery.}
\begin{center}
\includegraphics[width=\columnwidth]{images/map-crop.pdf}
\end{center}
\caption[\mapcap]{\mapcap}
\label{fig:map}
\end{figure}
%%%%%%%%%%%%%%%

@@

Although previous work using genome-scale data has shown that many of these populations or stocks can be resolved \cite{thompson2020complex,meek2020identifying},
such data is not practical for most large-scale applications of GSI and PBT, particularly because
of the much higher costs, longer turnaround time and challenges of producing highly
replicable datasets with reduced representation and low-coverage whole genome sequencing methods \cite{ali2016rad}. 
Accurate identification of the distinct ecotypes of Chinook salmon in the California Central Valley is of utmost importance for designing and implementing effective conservation and management actions. This has been challenging, given the recent common ancestors of these ecotypes and ongoing migration between subbasins. We describe the first set of genetic markers that identifies the late-fall-run salmon ecotype, which occurs only in the CCV and shares a genomic background with the more common CCV fall-run salmon ecotype. Moreover, we show how the FRH spring-run lineage is easily identifiable through a combination of traditional GSI and the characterization of functional genetic markers in the RoSA. Finally, although previous work has described GSI capabilities that distinguish the natural-spawning CCV spring-run lineages from each other and their fall-run counterparts with moderately high accuracy, we demonstrate near complete accuracy in distinguishing these 'stocks'. Moreover, the few fish that are apparently misidentified (Fig. 3) are likely to be primarily migrants and not true misidentifications. For example, the three fish that were field-characterized as spring-run from Butte Creek, but are genetically identified as winter-run salmon are clearly carry the winter-run genomic background and could not realistically be misidentified on the basis of the genetic data. This marker set and baseline reference dataset also provides unprecedented power for identifying fish from not only the other ESUs in California, but even the individual populations within those ESUs (Fig. 3).
Additionally, previous genetic marker sets for California Chinook salmon had sufficient power for inferring parent-offspring relationships, but only when both parents were sampled and genotyped. This marker set considerably increases the capacity for relationship inference in California salmon, by providing sufficient power for parent-offspring relationship inference, when only one parent was sampled, as well as identifying pairs of full siblings when no parents are sampled. Groups of full siblings larger than two are identified even more accurately than expressed in Figure 4.
